{RILS}-{ROLS}: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-12-13 paper_resources/doc/bmc_jbd/bmc_jbd.tex
Final proofread

Our method relies on the following operations: $+$, $-$, $\cdot$, $/$, $\sqrt{x}$, $x^2 $, $\sin$, $\cos$, $\log$, $\exp$, $a^x$, where $a^x$ is only consequently used -- it is never explicitly introduced in the search space. Beside these, the following set of constants enters the search space explicitly: $-1$, $0$, $1$, $2$, $\pi$, and $10$. 
Before we provide details of our method for solving SR, we will explain the two building blocks of \textsc{RILS-ROLS}: 1) iterated local search (ILS) and 2) ordinary least squares (OLS).

@@

\begin{algorithm}
     \textbf{Input}: solution $s$ \\
     \textbf{Output}: local perturbations (1-perturbations) of solution $s$ -- $s_{perturbations}$ \\
	\begin{algorithmic}[1] 
		\Procedure{All1Perturbations}{$s$}
		\State $s_{perturbations} \gets \emptyset$ \label{line:pertInit}
		\State $s \gets$ \texttt{NormalizeConstants}($s$)
		\State $s \gets$ \texttt{Simplify}($s$)
		\State $s_{subtrees} \gets$ \texttt{SubTrees}($s$)
		\For{$n \in s_{subtrees}$} \label{line:forSSStart}
		\State $n_{perturbations} \gets$ \texttt{All1PerturbationsAroundNode}($s$, $n$)
		\State $s_{perturbations} \gets s_{perturbations} \cup n_{perturbations}$	
		\EndFor \label{line:forSSEnd}
		\State \Return $s_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of all 1-perturbations of a given solution.}
	\label{alg:pert}
\end{algorithm}

@@

This is followed by constant normalization during which coefficients that enter multiplication, division, addition or subtraction are set to 1, while those entering the power function are rounded to integer, with the exception of square root, which is kept intact. For example, for expression $3.3\cdot(x+45.1\cdot y^{3.2})\cdot 81\cdot x/\sqrt{y}$ the normalized version is $1\cdot (x+1\cdot y^3)\cdot 1\cdot x/\sqrt{y}$. The reason for performing normalization is reducing the search space of possible perturbations. This reduction is reasonable, since normalization preserves the essential functional form. Note that coefficients get tuned later: the linear coefficient during the OLS phase, and the remaining ones during local search.

@@

\begin{algorithm}	
	\hspace*{\algorithmicindent} \textbf{Input}: solution $s$, node $n$\\
	\hspace*{\algorithmicindent} \textbf{Output}: 1-perturbations of solution $s$ around node $n$
	\begin{algorithmic}[1]
		\Procedure{All1PerturbationsAroundNode}{$s$, $n$}
		\State $n_{perturbations} \gets \emptyset$
		\If{$n = s$} \label{alg:alg4-case-1}
		\State $n_{perturbations} \gets n_{perturbations} \cup$ \texttt{NodeChanges}($n$)
		\EndIf
		\If{$n.arity \geq 1$}\label{alg:alg4-case-2}
		\State $n_{changes} \gets$ \texttt{NodeChanges}($n.left$)
		\For{$nc \in n_{changes}$}
		\State $new \gets$ \texttt{Replace}($s$, $n.left$, $nc$)
		\State $n_{perturbations} \gets n_{perturbations} \cup \{new\}$
		\EndFor
		\EndIf
		\If{$n.arity = 2$}\label{alg:alg4-case-3}
		\State $n_{changes} \gets$ \texttt{NodeChanges}($n.right$)
		\For{$nc \in n_{changes}$}
		\State $new \gets$ \texttt{Replace}($s$, $n.right$, $nc$)
		\State $n_{perturbations} \gets n_{perturbations} \cup \{new\}$
		\EndFor
		\EndIf
		\State \Return $n_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of 1-perturbations of a given solution around given node.}
	\label{alg:pertNode}
\end{algorithm}

@@

\emph{Case 1}. 
The observed node $n$ is the whole tree $s$ (see line ~\ref{alg:alg4-case-1} in Algorithm~\ref{alg:pertNode}). Based on  the previous exemplary expression tree, this means that the multiplication node that connects $(x+y^3)$ and $x/\sqrt{y}$ is to be changed. For example, multiplication can be replaced by addition, which forms a 1-perturbation expression (tree) $(x+y^3)+ x/\sqrt{y}$.

\emph{Case 2}. 
Node $n$ has the arity of at least 1 (see line ~\ref{alg:alg4-case-2} in Algorithm~\ref{alg:pertNode}). This means that the left subtree exists, so the left subtree node is to be changed. 
For example, if $n=x+y^3$, the overall perturbation might be $(x/y^3)+x/\sqrt{y}$ (addition is replaced by division). 
Another example would be the case of unary operation, e.g., when $n=\sqrt{y}$. In that case, some possible perturbations could be $(x+y^3)\cdot x/\sqrt{\ln{y}}$ (application of logarithm to the left subtree $y$) or $(x+y^3)\cdot x/\sqrt{x}$ (changing variable $y$ to $x$), etc.

@@

\begin{algorithm}
	\hspace*{\algorithmicindent} \textbf{Input}: perturbation $p$, sample training dataset $D_{tr}'$ \\
	\hspace*{\algorithmicindent} \textbf{Output}: local optimum $bp$ in the vicinity of perturbation $p$
	\begin{algorithmic}[1] 
		\Procedure{LocalSearch}{$p$, $D_{tr}'$}
		\State $bp, bp_{fit} \gets p,\ $\texttt{Fitness}$(p,D_{tr}')$ 
		\State $improved \gets true$
		\While{$improved$}
		\State $improved \gets false$
		\State $bp_{candidates} \gets \emptyset$
		\State $bp_{subtrees} \gets $\texttt{SubTrees}($bp$)
		\For{$n \in bp_{subtrees}$}
		\State $n_{candidates} \gets $ \texttt{All1PerturbationsAroundNodeExtended}($bp$, $n$)
		\For{$new \in n_{candidates}$}
		\State $new \gets$ \texttt{FitOLS}($new$, $D_{tr}'$)
		\State $new_{fit} \gets$ \texttt{Fitness}($new$, $D_{tr}'$)
		\If{$new_{fit} < bp_{fit}$}
		\State $bp, bp_{fit} \gets new, new_{fit}$
		\State $improved \gets true$
		\EndIf
		\EndFor
		\EndFor
		\EndWhile
		\State \Return $bp$
		\EndProcedure
	\end{algorithmic}
	\caption{Local search procedure.}
	\label{alg:ls}
\end{algorithm}

@@

\begin{itemize}
	\item Concerning the exact percentages in case when no noise is utilized in input data, the best performing algorithm is our \textsc{RILS-ROLS}. It found the exact solution in 57.84\% of runs for \textsc{Feynman} instances. It is even more impressive in solving \textsc{Strogatz} instances, when it was successful in 83.57\% runs. The second best performing  approach is \textsc{AI-Feynman}, successful in 55.78\% runs for the \textsc{Feynman} set, and 27.14\% runs for the problem instances from the benchmark set \textsc{Strogatz}. All other approaches perform significantly worse, and none of them are able to deliver the exact solution in more than 30\% runs over the instances of any of the two considered ground-truth benchmark sets. The overall results are in favor of \textsc{RILS-ROLS} with 60.62\%, while the second best is \textsc{AI-Feynman} with 52.65\%. 
	\item  Concerning the exact results in the presence of small noise (0.001), our \textsc{RILS-ROLS} still performs well, finding the exact model in 42.08\% of runs considering all problem instances. The second best approach is again \textsc{AI-Feynman}, successful in 31.89\% runs. As in the no-noise scenario, there is a significant difference in \textsc{AI-Feynman} performances among two sets of instances -- it performs much better on \textsc{Feynman} instances than on \textsc{Strogatz}. It is also worth mentioning that \textsc{AFP-FE} shows to be robust w.r.t. noise so its performances only slightly deteriorate in comparison to the no-noise scenario. This puts it on the third place with the overall accuracy of 21.23\%. The accuracy of other approaches is below 20\%.  
	\item  Concerning the scenario with high noise (level of 0.01), our \textsc{RILS-ROLS} still performs quite well -- it finds and exact model in 34.77\% runs. The second best is \textsc{AFP-FE}, reaching 20\% correct solutions. \textsc{AI-Feynman} performances degrade significantly under this level of noise, so it is the 6-th overall ranking, reaching the accuracy of only 12.61\%.

@@

In addition to well-known benchmarks from the literature, we introduced a new non-biased benchmark set of randomly generated instances, labeled by \textsc{Random}. 
This benchmark was used to test \textsc{RILS}-\textsc{ROLS} scalability, by varying the size of the model and the number of variables.

>>>>

Our method relies on the following operations: $+$, $-$, $\cdot$, $/$, $\sqrt{x}$, $x^2 $, $\sin$, $\cos$, $\log$, $\exp$, $a^x$, where $a^x$ is only consequently used -- it is never explicitly introduced in the search space. Beside these, the following set of constants enters the search space explicitly: $-1$, $0$, $1$, $2$, $\pi$, and $10$. 
Before we provide details of our method for solving SR, we will explain the two building blocks of \textsc{RILS-ROLS}: 1) iterated local search (ILS) and 2) ordinary least square (OLS).

@@

\begin{algorithm}

	\begin{algorithmic}[1] 
		\Statex   \textbf{Input}: solution $s$ 
		\Statex \textbf{Output}: local perturbations (1-perturbations) of solution $s$ -- $s_{perturbations}$
		\Procedure{All1Perturbations}{$s$}
		\State $s_{perturbations} \gets \emptyset$ \label{line:pertInit}
		\State $s \gets$ \texttt{NormalizeConstants}($s$)
		\State $s \gets$ \texttt{Simplify}($s$)
		\State $s_{subtrees} \gets$ \texttt{SubTrees}($s$)
		\For{$n \in s_{subtrees}$} \label{line:forSSStart}
		\State $n_{perturbations} \gets$ \texttt{All1PerturbationsAroundNode}($s$, $n$)
		\State $s_{perturbations} \gets s_{perturbations} \cup n_{perturbations}$	
		\EndFor \label{line:forSSEnd}
		\State \Return $s_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of all 1-perturbations of a given solution.}
	\label{alg:pert}
\end{algorithm}

@@

This is followed by constant normalization during which coefficients that enter multiplication, division, addition or subtraction are set to 1, while those entering the power function are rounded to integer, with the exception of square root, which is kept intact. For example, for expression $3.3\cdot(x+45.1\cdot y^{3.2})\cdot 81\cdot x/\sqrt{y}$ the normalized version is $1\cdot (x+1\cdot y^3)\cdot 1\cdot x/\sqrt{y}$. The reason for performing normalization is reducing the search space of possible perturbations. This reduction is reasonable, since normalization preserves the essential functional form. Note that coefficients get tuned later: the linear coefficient during the OLS phase, and the remaining ones during local search, see Section~\ref{sec:ls}.

@@

\begin{algorithm}

	\begin{algorithmic}[1]
		\Statex  \textbf{Input}: solution $s$, node $n$ 
		\Statex  \textbf{Output}: 1-perturbations of solution $s$ around node $n$
		\Procedure{All1PerturbationsAroundNode}{$s$, $n$}
		\State $n_{perturbations} \gets \emptyset$
		\If{$n = s$} \label{alg:alg4-case-1}
		\State $n_{perturbations} \gets n_{perturbations} \cup$ \texttt{NodeChanges}($n$)
		\EndIf
		\If{$n.arity \geq 1$}\label{alg:alg4-case-2}
		\State $n_{changes} \gets$ \texttt{NodeChanges}($n.left$)
		\For{$nc \in n_{changes}$}
		\State $new \gets$ \texttt{Replace}($s$, $n.left$, $nc$)
		\State $n_{perturbations} \gets n_{perturbations} \cup \{new\}$
		\EndFor
		\EndIf
		\If{$n.arity = 2$}\label{alg:alg4-case-3}
		\State $n_{changes} \gets$ \texttt{NodeChanges}($n.right$)
		\For{$nc \in n_{changes}$}
		\State $new \gets$ \texttt{Replace}($s$, $n.right$, $nc$)
		\State $n_{perturbations} \gets n_{perturbations} \cup \{new\}$
		\EndFor
		\EndIf
		\State \Return $n_{perturbations}$
		\EndProcedure
	\end{algorithmic}
	\caption{Generation of 1-perturbations of a given solution around given node.}
	\label{alg:pertNode}
\end{algorithm}

@@

\emph{Case 1}. 
The observed node $n$ is the whole tree $s$ (see line~\ref{alg:alg4-case-1} in Algorithm~\ref{alg:pertNode}). Based on  the previous exemplary expression tree, this means that the multiplication node that connects $(x+y^3)$ and $x/\sqrt{y}$ is to be changed. For example, multiplication can be replaced by addition, which forms a 1-perturbation expression (tree) $(x+y^3)+ x/\sqrt{y}$.

\emph{Case 2}. 
Node $n$ has the arity of at least 1 (see line~\ref{alg:alg4-case-2} in Algorithm~\ref{alg:pertNode}). This means that the left subtree exists, so the left subtree node is to be changed. 
For example, if $n=x+y^3$, the overall perturbation might be $(x/y^3)+x/\sqrt{y}$ (addition is replaced by division). 
Another example would be the case of unary operation, e.g., when $n=\sqrt{y}$. In that case, some possible perturbations could be $(x+y^3)\cdot x/\sqrt{\ln{y}}$ (application of logarithm to the left subtree $y$) or $(x+y^3)\cdot x/\sqrt{x}$ (changing variable $y$ to $x$), etc.

@@

\begin{algorithm}

	\begin{algorithmic}[1] 
		\Statex \textbf{Input}: perturbation $p$, sample training dataset $D_{tr}'$   
		\Statex \textbf{Output}: local optimum $bp$ in the vicinity of perturbation $p$
		\Procedure{LocalSearch}{$p$, $D_{tr}'$}
		\State $bp, bp_{fit} \gets p,\ $\texttt{Fitness}$(p,D_{tr}')$ 
		\State $improved \gets true$
		\While{$improved$}
		\State $improved \gets false$
		\State $bp_{candidates} \gets \emptyset$
		\State $bp_{subtrees} \gets $\texttt{SubTrees}($bp$)
		\For{$n \in bp_{subtrees}$}
		\State $n_{candidates} \gets $ \texttt{All1PerturbationsAroundNodeExtended}($bp$, $n$)
		\For{$new \in n_{candidates}$}
		\State $new \gets$ \texttt{FitOLS}($new$, $D_{tr}'$)
		\State $new_{fit} \gets$ \texttt{Fitness}($new$, $D_{tr}'$)
		\If{$new_{fit} < bp_{fit}$}
		\State $bp, bp_{fit} \gets new, new_{fit}$
		\State $improved \gets true$
		\EndIf
		\EndFor
		\EndFor
		\EndWhile
		\State \Return $bp$
		\EndProcedure
	\end{algorithmic}
	\caption{Local search procedure.}
	\label{alg:ls}
\end{algorithm}

@@

\begin{itemize}
	\item Concerning the exact percentages in case when no noise is utilized in input data, the best performing algorithm is our \textsc{RILS-ROLS}. It found the exact solution in 57.84\% of runs for \textsc{Feynman} instances. It is even more impressive in solving \textsc{Strogatz} instances, as it was successful in 83.57\% runs. The second best performing  approach is \textsc{AI-Feynman}, successful in 55.78\% runs for the \textsc{Feynman} set, and 27.14\% runs for the problem instances from the benchmark set \textsc{Strogatz}. All other approaches perform significantly worse, and none of them are able to deliver the exact solution in more than 30\% runs over the instances of any of the two considered ground-truth benchmark sets. The overall results are in favor of \textsc{RILS-ROLS} with 60.62\%, while the second best is \textsc{AI-Feynman} with 52.65\%. 
	\item  Concerning the exact results in the presence of small noise (0.001), our \textsc{RILS-ROLS} still performs well, finding the exact model in 42.08\% of runs considering all problem instances. The second best approach is again \textsc{AI-Feynman}, successful in 31.89\% runs. As in the no-noise scenario, there is a significant difference in \textsc{AI-Feynman} performances among two sets of instances -- it performs much better on \textsc{Feynman} instances than on \textsc{Strogatz}. It is also worth mentioning that \textsc{AFP-FE} shows to be robust w.r.t. noise so its performances only slightly deteriorate in comparison to the no-noise scenario. This puts it on the third place with the overall accuracy of 21.23\%. The accuracy of other approaches is below 20\%.  
	\item  Concerning the scenario with high noise (level of 0.01), our \textsc{RILS-ROLS} still performs quite well -- it finds and exact model in 34.77\% runs. The second best is \textsc{AFP-FE}, reaching 20\% correct solutions. \textsc{AI-Feynman} performances degrade significantly under this level of noise, so it is the 6-th overall ranking, reaching the accuracy of only 12.61\%.

@@

In addition to well-known benchmarks from the literature, we introduced a new non-biased benchmark set of randomly generated instances.  %, labeled by \textsc{Random}. 
This benchmark was used to test \textsc{RILS}-\textsc{ROLS} scalability, by varying the size of the model and the number of variables.
